Likelihoods for a general class of ARGs under the SMC

### 2025-02-21 paper.tex
Add results for tsinfer+tsdate

% % % % % % % % % % % % % % % % % %
\subsection*{Robustness to partial information}
Eq.~\eqref{eq:full-lik} allows us to compute a likelihood for \emph{any}
collection of nodes (ancestral haplotypes that lived at some time 
$t$ in the past) and edges (a record that child node $c$ 
inherited the genomic interval $[x, y)$ from parent node $p$)
under the SMC. Other than some basic consistency constraints (e.g., parents
must be older than children) there are no requirements on the topology
described, and in particular, no requirement that the events from the 
model (the SMC) should be explicitly identified in the graph. Our 
formulation reasons about the events that \emph{could have resulted in}
the observed nodes and edges, and therefore frees us from the requirement
of having to explicitly write those down. In this section we 
illustrate the robustness
of this approach to incomplete information about recombination and 
coalescent events via a proof-of-concept example, in which we 
remove various levels of detail from a simulated ARG.

\begin{figure}
    \centering
    \includegraphics[width=0.8\textwidth]{figures/likelihood_surface}
    \caption{Log likelihood-ratio curves for ARGs containing
     varying degrees of information about the realised simulation.
    The ``Full ARG'' here is the result of an 
    \texttt{msprime} simulation of the SMC (100 diploid samples, genome length 1Mb, 
    $r=2.5\times10^{-8}$, $N_e=1\times10^4$) which contains complete 
    information about the simulated events 
    (9,378 nodes; 13,221 edges; 2,933 trees). 
    The ``RE nodes removed'' ARG is the 
    result of simplifying out all of the recombination nodes (3,332 nodes;
    6,901 edges; note that \texttt{msprime} uses distinct nodes for the 
    left and right parent of each recombination). The ``Fully simplified''
    ARG has all locally unary nodes removed from trees (3,332 nodes; 10,674
    edges). Each of these ARGs additionally has 5\% and 25\% of the 
    internal nodes removed to create polytomies.
    See the text for details on how the likelihood ratio values are computed.
    The ``Full ARG'' and ``RE nodes deleted'' lines coincide.
    \label{fig:lik-surface}}
\end{figure}

In Fig~\ref{fig:lik-surface} we begin with a realisation of the SMC as a ``full
ARG'', that is, containing complete information about every simulated event.
We then plot the likelihood computed by Eq.~\eqref{eq:full-lik} 
as we vary the recombination rate parameter $r$ (as a ratio
of the likelihood computed at the true value of $r$). Unsurprisingly, we can
see that the minimum of the likelihood-ratio curve when using the 
full simulated ARG is at the true value of the parameter.
What is surprising is that we obtain an indistinguishable curve when 
we remove all the recombination nodes from this ARG, and therefore 
lose information about the precise timing of recombination events.
In this case, the locally unary portions of coalescent nodes have given
all the information that we need about recombination events.
Removing all locally unary nodes in the fully simplified ARG does result
in a loss of information, and we can see that the minimum of the 
likelihood ratio surface is biased away from its true value.

Fig~\ref{fig:lik-surface} shows that Eq.~\eqref{eq:full-lik} is also remarkably
robust to the presents of polytomies in this example. For each of the three
ARGs discussed in the previous paragraph, we also evaluated the likelihood
for ARGs in which 5\% or 25\% of the internal ARG nodes are removed and 
edges adjusted accordingly.
This introduces polytomies at varying degrees in the different ARGs because 
deleting a node in a tree (and connecting its children directly to its
parent) will only create a polytomy if the parent 
is not unary. Thus, while deleting 25\% of the internal nodes 
in the fully simplified ARG created an average of 33 polytomies
over the 2,933 trees along the sequence (each tree having 
200 leaf nodes), it resulted in an average of 21 polytomies
per tree in the ARG where recombination nodes have been removed,
and only 10 in the full ARG (which has an average of 1026 unary 
nodes per tree).
%                     name    leaves  unary       binary      poly
% 0   poly_full_arg_smc_0.05  200.0   1025.565973 192.662462  3.153086
% 1   poly_unary_coal_0.05    200.0   429.315017  190.182253  4.372696
% 2   poly_fully_sfiied_0.05  200.0   0.000000    185.691468  6.393174
% 3   poly_full_arg_smc_0.25  200.0   779.846233  176.830549  10.307876
% 4   poly_unary_coal_0.25    200.0   323.054926  154.422588  21.104016
% 5   poly_fully_sified_0.25  200.0   0.000000    117.096174  33.241986
Given this disparity in the number of polytomies introduced, the 
effects on the likelihood across the three examples are not entirely
comparable. It is remarkable, however, how little effect introducing 
these polytomies has on the likelihood surface in this example. For 
the ARGs containing unary nodes, removing 25\% of the internal 
nodes has negligible effect, and even on the fully simplified 
ARG has a very minor effect on the minimum.

@@

The key contribution of this paper is to rederive a classical
ARG likelihood function in a way that supports the sort of partial information
provided by some modern ARG inference tools,
using a novel backwards-time derivation of the SMC.
The underlying process that we reason about probabilistically is still
composed of events affecting lineages, but, because we
do not necessarily observe or estimate these events,
we integrate over the possible timings implied by the information
in the input ARG.
Our method can compute a likelihood under the SMC for a very general class of 
ARG~\citep{wong_general_2023}, 
a significant increase in flexibility over the 
strict input requirements for the classical Kuhner-Yamato-Felsenstein 
formulation. 
We have shown via some illustrative examples that the formulation is 
remarkably robust to incomplete information, tolerating the loss 
of a significant fraction of the internal nodes in the ARG to minimal effect.

Although we can associate a likelihood with any ARG in \texttt{tskit} format,
there are some caveats. By treating each edge independently,
our algorithm will still produce a value for ARGs with
the various degrees of completeness to which inheritance
is documented by the different ARG inference methods.
% : e.g., polytomies, absence
% of locally unary coalescence nodes, unpreserved identify of internal nodes
% across local trees, etc.
% However, many of these imprecisions violate the implicit
% assumptions of the likelihood computation.
For example, we assume the 
presence of locally unary nodes to correctly identify the
full extent of all lineages going back through time 
and therefore recombination events, but this information
may be incomplete (e.g., the full simplifed ARG in Fig.~\ref{fig:lik-surface})
or even incorrect.
% Without these unary nodes, our algorithm will detect a recombination
% event along both coalescing lineages in case of a (sub)tree height changes along
% the ARG as both edges would switch parent nodes. 
Locally unary nodes are not a common aspect of current ARG 
research~\citep{wong_general_2023}, but 
are estimated by tsinfer, and can be imputed to high
degree of accuracy from fully simplified ARGs~\citep{fritze2024forest}.
Determining the effects of various forms of \emph{incorrect} topology on
these likelihoods requires further research.

>>>>

% % % % % % % % % % % % % % % % % %
\subsection*{Robustness to partial and incorrect information}
Eq.~\eqref{eq:full-lik} allows us to compute a likelihood for \emph{any}
collection of nodes (ancestral haplotypes that lived at some time 
$t$ in the past) and edges (a record that child node $c$ 
inherited the genomic interval $[x, y)$ from parent node $p$)
under the SMC. Other than some basic consistency constraints (e.g., parents
must be older than children) there are no requirements on the topology
described, and in particular, no requirement that the events from the 
model (the SMC) should be explicitly identified in the graph. Our 
formulation reasons about the events that \emph{could have resulted in}
the observed nodes and edges, and therefore frees us from the requirement
of having to explicitly write those down. In this section we 
illustrate the robustness
of this approach to incomplete and potentially incorrect
information about recombination and 
coalescent events via a proof-of-concept example. 
% in which we 
% remove various levels of detail from a simulated ARG.

\begin{figure}
    \centering
    \includegraphics[width=0.8\textwidth]{figures/likelihood_surface}
    \caption{Log likelihood-ratio curves for ARGs with various properties.
    The ``Full ARG'' here is the result of an 
    msprime simulation of the SMC (100 diploid samples, genome length 1Mb, 
    $r=\mu=2.5\times10^{-8}$, $N_e=1\times10^4$).
    The ``RE nodes removed'' ARG is the 
    result of simplifying out all of the recombination nodes,
    and the ``Fully simplified'' ARG has all locally unary nodes removed. 
    Each of these simulated ARGs additionally has 5\% and 25\% of the 
    internal nodes removed to create polytomies.
    We also show the results for ARGs inferred from the original simulation
    data using tsinfer and tsdate 
    (see text for details), in which we use the default output of tsinfer
    including unary nodes (tsinfer+tsdate), and in which we fully simplify 
    before passing to tsdate (tsinfer+tsdate FS). 
    See the text for details on how the likelihood ratio 
    values are computed.
    The ``Full ARG'' and ``RE nodes removed'' lines coincide.
    \label{fig:lik-surface}}
\end{figure}

In Fig~\ref{fig:lik-surface} we begin with a realisation of the SMC 
containing complete information about every simulated event
(Full ARG; 9,378 nodes; 13,221 edges; 2,933 trees). 
We then plot the likelihood computed by Eq.~\eqref{eq:full-lik} 
as we vary the recombination rate parameter $r$, as a ratio
of the likelihood at the true value of $r$. 
Reassuringly, we can
see that the minimum of the likelihood-ratio curve when using the 
full simulated ARG is close to the true value of the parameter.
(The likelihood ratio curve obtained when computing the likelihood
under the full coalescent with recombination model using the KYF
formula is identical; not shown.)
We next compute the likelihood ratio for the ARG obtained when we 
simplify out the recombination nodes from the full ARG
(RE nodes removed; 3,332 nodes; 6,901 edges; 2,933 trees; note that msprime 
uses distinct nodes for the left and right parent of each recombination).
The curve we obtain is indistinguishable from that of the full
ARG, demonstrating that the locally unary portions of coalescent nodes 
contain all the information that we need about recombination events
for this application.
When we fully simplify the simulated ARG
to remove all locally unary nodes from trees 
(Fully simplified; 3,332 nodes; 10,674 edges, 2,933 trees) 
we can see there is a significant loss of information,
and the minimum of the 
likelihood ratio surface is biased away from its true value.
(See~\cite{wong_general_2023} for more discussion on varying 
degrees of ARG simplification.)

Fig~\ref{fig:lik-surface} shows that Eq.~\eqref{eq:full-lik} is also remarkably
robust to the presences of polytomies in this example. For each of the three
ARGs discussed in the previous paragraph, we also evaluated the likelihood
for ARGs in which 5\% or 25\% of the internal ARG nodes are removed and 
edges adjusted accordingly.
This introduces polytomies at varying degrees in the different ARGs because 
deleting a node in a tree (and connecting its children directly to its
parent) will only create a polytomy if the parent 
is not unary. Thus, while deleting 25\% of the internal nodes 
in the fully simplified ARG created an average of 33 polytomies
over the 2,933 trees along the sequence (each tree having 
200 leaf nodes), it resulted in an average of 21 polytomies
per tree in the ARG where recombination nodes have been removed,
and only 10 in the full ARG (which has an average of 1026 unary 
nodes per tree).
%     name                  leaves  unary       binary  poly
% 0   full_arg_smc          200.0   1094.417661 199.000000  0.000000
% 1   unary_coal            200.0   456.396522  199.000000  0.000000
% 2   fully_simplified      200.0   0.000000    199.000000  0.000000
% 3   tsinfer_tsdate        200.0   192.447590  119.754399  32.270849
% 4   tsinfer_fs_tsdate     200.0   0.000000    119.563898  32.322684
% 5   py_full_arg_smc_0.05  200.0   1025.565973 192.662462  3.153086
% 6   py_unary_coal_0.05    200.0   429.315017  190.182253  4.372696
% 7   py_fully_simied_0.05  200.0   0.000000    185.691468  6.393174
% 8   py_full_arg_smc_0.25  200.0   779.846233  176.830549  10.307876
% 9   py_unary_coal_0.25    200.0   323.054926  154.422588  21.104016
% 10  py_fully_simied_0.25  200.0   0.000000    117.096174  33.241986
Given this disparity in the number of polytomies introduced, the 
effects on the likelihood across the three examples are not entirely
comparable. For the ARGs containing unary nodes, 
removing 25\% of the internal 
nodes has negligible effect, and even on the fully simplified 
ARG has a very minor effect on location of the minimum.

Fig~\ref{fig:lik-surface} also shows results for ARGs inferred from the 
simulated data using tsinfer and tsdate~\citep{wohns_unified_2022},
using default parameters for tsinfer, and the true values of mutation
rate and $N_e$ for tsdate.
In the first, we use the default output of tsinfer 
(2,148 nodes; 5,589 edges; 1,307 trees; 3,039 mutations).
This ARG has an average of 192 unary nodes, 120 binary nodes, and 32 nodes
with three or more children per tree. Tsinfer does not directly estimate
recombination events; the breakpoints between trees are the consequence of
switches between parental haplotypes in the 
\cite{li2003modeling} copying process. Information about 
recombination events in the returned ARG is therefore quite imprecise,
and likely to contain incorrect and contradictory details if one were to 
attempt to reconstruct the exact subtree-prune-and-regraft 
moves~\citep[e.g.][]{rasmussen2023espalier}. Nonetheless, in
this toy example at least, our likelihood function performs well
on the output of tsinfer, with the minimum close to the true
parameter value. We also show the likelihood curve obtained when we 
fully simplify the output of tsinfer 
(1,865 nodes; 7,511 edges; 1,252 trees; 3,039 mutations)
before dating. We can see that, as with the case of the true data,
removing all the unary nodes results in a significant bias
away from the true value of the recombination rate parameter.

@@

The key contribution of this paper is to rederive a classical
ARG likelihood function in a way that supports the sort of partial information
provided by some modern ARG inference tools,
using a novel backwards-time derivation of the SMC.
The underlying process that we reason about probabilistically is still
composed of events affecting lineages, but, because we
do not necessarily observe or estimate these events,
we integrate over the possible timings implied by the information
in the input ARG.
Our method can compute a likelihood under the SMC for a very general class of 
ARG~\citep{wong_general_2023}, 
a significant increase in flexibility over the 
strict input requirements for the classical Kuhner-Yamato-Felsenstein 
formulation. 
We have shown via some illustrative examples that the formulation is 
remarkably robust to incomplete information, tolerating the loss 
of a significant fraction of the internal nodes in the ARG to minimal effect.
% I think we have to grasp this nettle or we're inviting the reviewers to 
% give us a bunch of "what if you tried this" bits of extra homework.
While there are many different potential applications to concrete 
inferential problems, the details are nontrivial and best consigned to future work.

A major advantage of our approach is that it does not require
us to explicitly identify common ancestor and recombination events 
in the input. Indeed, 
we can compute a likelihood for any ARG in \texttt{tskit} format,
potentially bridging the gap between population genetic theory and the 
varied structures inferred by current methods~\citep{wong_general_2023}.

there are some caveats. By treating each edge independently,
our algorithm will still produce a value for ARGs with
the various degrees of completeness to which inheritance
is documented by the different ARG inference methods.
% : e.g., polytomies, absence
% of locally unary coalescence nodes, unpreserved identify of internal nodes
% across local trees, etc.
% However, many of these imprecisions violate the implicit
% assumptions of the likelihood computation.
For example, we assume the 
presence of locally unary nodes to correctly identify the
full extent of all lineages going back through time 
and therefore recombination events, but this information
may be incomplete (e.g., the full simplifed ARG in Fig.~\ref{fig:lik-surface})
or even incorrect.
% Without these unary nodes, our algorithm will detect a recombination
% event along both coalescing lineages in case of a (sub)tree height changes along
% the ARG as both edges would switch parent nodes. 
Locally unary nodes are not a common aspect of current ARG 
research~\citep{wong_general_2023}, but 
are estimated by tsinfer, and can be imputed to high
degree of accuracy from fully simplified ARGs~\citep{fritze2024forest}.
Determining the effects of various forms of \emph{incorrect} topology on
these likelihoods requires further research.
